---
title: "JSC370 Final Project"
author: "Dennis Tat"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---


---
title: "Income vs. Inflation: Understanding
Housing Affordability in Canada Across
Generations"
output: html_document
date: "2025-04-25"
---

```{r setup, include=FALSE}
set.seed(1007)
library(tidyr)
library(dplyr)
library(ggplot2)
library(microbenchmark)
library(parallel)
library(rpart)
library(rpart.plot)
library(randomForest)
library(kableExtra)
library(tidyverse)
library(plotly)
library(widgetframe)
library(tidytext)
library(mgcv)
library(caret)
library(knitr)

suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))

income <- read.csv("Data/income.csv")
cpi <- read.csv("Data/cpi.csv")
housing <- read.csv("Data/housing.csv")
```


## Introduction

With the next federal election just around the corner, issues of affordability
and housing access have taken center stage in public discourse. Canada has
experienced a notable increase in the cost of living over the past decade, evident
by the rising inflation. These economic pressures have outpaced the growth
of salaries, making it increasingly difficult for many Canadian young adults
entering the workforce to achieve financial stability. For this demographic, home
ownership has become an increasingly distant prospect. At the same time,
those who already own property have seen their real estate assets appreciate
significantly in value, contributing to a growing sense of inequality between
generations.

This widening gap has fueled a prevailing narrative that older generations had
a far easier path to financial security and home ownership than today’s youth.
While anecdotal evidence and media commentary have popularized this belief,
the goal of this study is to perform empirical analysis to determine the extent
to which this perception is grounded in economic reality.

This study seeks to explore these dynamics by examining the relationship be-
tween inflation, housing prices, and salaries, in order to provide a clearer picture
of how economic conditions have evolved over time.

## Research Questions 

To investigate the aforementioned imbalance in spending power between younger
and older generations, this study will evaluate a multitude of economic factors.
Several key questions will guide the analysis:

Has the median income of Canadians kept pace with inflation over time?

How has median income evolved across age groups, especially for younger versus
older Canadians?

How does the change in income compare to the rise in housing prices?

Finally, can housing prices be predicted by analyzing the price trends of other
consumer goods?

This paper can be downloaded <a href="JSC370 Project Report.pdf" target="_blank">here</a>.



```{r echo=FALSE, include=FALSE}
# Cleaning Data Sets

housing <- housing |>
  mutate(Y = 100 / Value[Year == 2002],
         Adjusted.value = Value * Y) |>
  select(-Y)

income <- income %>%
  group_by(Age.group, Sex) %>%  
  mutate(Adjusted.value = Value / Value[Year == 2022])

income <- income %>%
  group_by(Age.group, Sex) %>%
  mutate(Y = 100 / Adjusted.value[Year == 2002], 
         Adjusted.value = Adjusted.value * Y) %>% 
  ungroup() %>%
  select(-Y)
```




```{r echo=FALSE, include=FALSE}
# EDA

unique(income$Age.group)
unique(income$Sex)
summary(income)
summary(cpi)
```

```{r echo=FALSE, include=FALSE}
# EDA Summary

kable(head(income[order(-income$Adjusted.value), ]), caption = "Largest Income Indexes", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)

kable(tail(income[order(-income$Adjusted.value), ]), caption = "Lowest Income Indexes", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)
```



```{r echo=FALSE, include=FALSE}
# EDA Summary

kable(head(cpi[order(cpi$Value), ]), caption = "Lowest CPI Values", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)

kable(tail(cpi[order(cpi$Value), ]), caption = "Largest CPI Values", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)
```


```{r echo=FALSE, include=FALSE}
# EDA Summary

cpi_filtered <- cpi %>%
  filter(Products.and.product.groups == "All-items") %>%
  filter(Year %in% c(1976, 2024))

kable(cpi_filtered, caption = "", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)

```



```{r echo=FALSE, include=FALSE}
# EDA Summary

cpi_filtered <- cpi %>%
  filter(Products.and.product.groups == "Eggs") %>%
  filter(Year %in% c(1976, 2024))

kable(cpi_filtered, caption = "", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)
```


```{r echo=FALSE, include=FALSE}
# EDA Summary

kable(head(housing[order(housing$Adjusted.value), ]), caption = "Largest Housing Index Values", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)

kable(tail(housing[order(housing$Adjusted.value), ]), caption = "Lowest Housing Index Values", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)
```


```{r echo=FALSE, warning=FALSE, include=FALSE}
# Graphing Consumer Pricing and Income Index Trends

both_sex <- income |>
  filter(Sex == "Both sexes") |>
  filter(Age.group == "25 to 54 years")

cpi_average <- cpi |>filter(Products.and.product.groups == "All-items") |> filter(Year >= 1976)

question1 <- merge(
  cpi_average,
  both_sex,
  by = c("Year")
)
ggplot(question1, aes(x = Year)) +
  geom_line(aes(y = Value.x, color = "CPI"), size = 1) +
  geom_line(aes(y = Adjusted.value, color = "Income Index"), size = 1, alpha = 0.7) +
  labs(title = "Consumer Pricing and Income Index Trends",
       x = "Year",
       y = "Index") +
  theme_minimal() +
  theme(legend.position = "right")
```


```{r echo=FALSE, include=FALSE}
# Graphing Income Index Trends by Age Group

both_sex_all_groups <- income |>
  filter(Sex == "Both sexes") |>
  filter(Age.group != "15 years and over")|>
  filter(Age.group != "25 to 54 years")
  
ggplot(both_sex_all_groups, aes(x = Year, y = Adjusted.value, color = Age.group)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(title = "Income Index Trends by Age Group",
       x = "Year", 
       y = "Index",
       color = "Age Group") +
  theme_minimal() + 
  theme(legend.position = "bottom") 
```


```{r echo=FALSE, include=FALSE}
# Graphing Income Index Trends by Sex

male_female_income <- income |>
  filter(Sex != "Both sexes") |>
  filter(Age.group == "25 to 54 years")

ggplot(male_female_income, aes(x = Year, y = Adjusted.value, color = Sex)) +
  geom_line(size = 1) + 
  geom_point(size = 3) +
  labs(title = "Income Index Trends by Sex",
       x = "Year", 
       y = "Index",
       color = "Sex") +
  theme_minimal() + 
  theme(legend.position = "right")

```



```{r echo=FALSE, include=FALSE}
# Graphing Housing and Income Index Trends

question3 <- merge(
  both_sex,
  housing,
  by = c("Year")
)

ggplot(question3, aes(x = Year)) +
  geom_line(aes(y = Adjusted.value.y, color = "Housing Index"), size = 1) +
  geom_line(aes(y = Adjusted.value.x, color = "Income Index"), size = 1, alpha = 0.7) +
  labs(title = "Housing and Income Index Trends",
       x = "Year",
       y = "Index") +
  theme_minimal() +
  theme(legend.position = "right")
```




```{r echo=FALSE, include=FALSE}
# Building Data Frame for Modeling

cpi_reshape <- cpi %>%
  pivot_wider(
    names_from = Products.and.product.groups,  
    values_from = Value
  )
```

```{r echo=FALSE, include=FALSE}
# Making variable names friendly for GAM

cpi_reshape_clean <- cpi_reshape |>
  select(Year, Food, Gasoline, "Mortgage interest cost", "Health care", "Dental care services",
         "Education", "Property taxes and other special charges", "Transportation")

cpi_reshape_clean <- merge(cpi_reshape_clean, housing, by = "Year", all.x = TRUE)

cpi_reshape_clean <- cpi_reshape_clean %>%
  rename_with(~ gsub(" ", ".", .), everything())
```


```{r echo=FALSE, include=FALSE}
# Linear Model

model <- lm(Adjusted.value ~ Food + `Mortgage.interest.cost`  +
            `Health.care` + `Property.taxes.and.other.special.charges` + Year + Gasoline +
              Transportation
            , data = cpi_reshape_clean)


summary(model)
```


```{r echo=FALSE, include=FALSE}
# GAM Model

gam_model <- gam(Adjusted.value ~ s(Year) + Food + `Mortgage.interest.cost`  +
            `Health.care` + `Property.taxes.and.other.special.charges` + Gasoline +
              Transportation, data = cpi_reshape_clean)

summary(gam_model)

```

```{r echo=FALSE, include=FALSE}
# Bagging Model

project_train_index <- createDataPartition(cpi_reshape_clean$Adjusted.value, p = 0.7, list = FALSE)
project_train <- cpi_reshape_clean[project_train_index, ]
project_test <- cpi_reshape_clean[-project_train_index, ]



tree <- rpart(Adjusted.value ~ ., 
                    data = project_train, 
                    method = "class",
                    control = rpart.control(minsplit = 10, 
                                            minbucket = 3, 
                                            cp = 0.01, 
                                            xval = 10))

plotcp(tree)
printcp(tree)

optimalcp = tree$cptable[which.min(tree$cptable[, "xerror"]), "CP"]

tree_prune <- prune(tree, cp = optimalcp)

bag<- randomForest(Adjusted.value~ ., 
                         data = project_train, 
                         mtry = ncol(project_train) - 1, 
                         importance = TRUE,
                         na.action = na.omit)
# oob error rate
bag_oob <- bag$mse[length(bag$mse)]
bag_oob

importance(bag)
```

```{r echo=FALSE, include=FALSE}
# Random Forest Model

rf <- randomForest(Adjusted.value ~ ., 
                         data = project_train, 
                         importance = TRUE, 
                         na.action = na.omit)
#oob error rate
rf_oob <- rf$mse[length(rf$mse)]
rf_oob

importance(rf)


predictions <- predict(model)

residuals <- cpi_reshape_clean$Adjusted.value - predictions

MSE <- mean(residuals^2)

gam_predictions <- predict(gam_model)

gam_residuals <- cpi_reshape_clean$Adjusted.value - gam_predictions

gam_MSE <- mean(gam_residuals^2)
```


```{r echo=FALSE, include=FALSE}
# Model Evaluation Summary

model_performance <- data.frame(
  Model = c(
    "Linear",
    "GAM",
    "Bagging",
    "Random Forest"
  ),
  Error_MSE = c(
    MSE,
    gam_MSE,
    bag_oob,
    rf_oob
  )
)

kable(model_performance, col.names = c("Model", "Error / MSE"), caption = "Model Performance Comparison")
```



## Interactive Visuals {.tabset}

### Historic Housing Price Index

```{r echo = FALSE}
cpi_reshape_clean |>
  plot_ly(
    x = ~Year,
    y = ~Adjusted.value,
    type = "scatter",
    mode = "markers",
    hoverinfo = "text",
    text = ~paste0("Housing Price Index: ", round(Adjusted.value, 2), "\nYear: ", Year)
  ) |>
  layout(
    title = "Housing Price Index through the years",
    xaxis = list(title = "Year") ,
    yaxis = list(title = "Housing Price Index"),
    hovermode = "compare"
  )
```

Housing prices have continued to increase since the 2010's. Prices reached a peak in 2022, 
but are still much higher than expected. 

### Historic Housing and Income Index

```{r echo=FALSE}
question3 |>
  plot_ly(
    x = ~Year,
    y = ~`Adjusted.value.x`,
    z = ~`Adjusted.value.y`,
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 5),
    hoverinfo = "text",
    text = ~paste0("Year: ", Year, "\nHousing Price Index: ", round(Adjusted.value.y), "\nIncome Index: ", round(Adjusted.value.x))
  ) |>
  layout(
    title = "3D scatterplot of Income Index and Housing Price Index",
    scene = list(
      xaxis = list(title = "Year") ,
      yaxis = list(title = "Income Index"),
      zaxis = list(title = "Housing Price Index")
    )
  )

```

Based on this plot, we see that the income index growth is much smaller than that of 
the housing price index. As such, it is becoming increasingly difficult for Canadians 
to purchase a home.

### Extreme Price Indexes

```{r echo=FALSE}

top_bottom <- cpi %>%
  unite("Name", Products.and.product.groups, Year, sep = " ")|>
  arrange(desc(Value)) %>%
  slice(c(1:5, (n() - 4):n()))  # top 5 and bottom 5



pl <- top_bottom|>
  ggplot(aes(x = fct_reorder(Name, Value), y = Value)) +
  geom_col() +
  coord_flip() +
  theme_bw()

ggplotly(pl)



```

Plotting the largest and smallest price indexes from our data set, we see that
the cost of cigarettes and tobacco products were extremely cheap in the 1970s, whereas 
things like gas nad fuel have skyrocketed in price. 
