---
title: "JSC370 Final Project"
author: "Dennis Tat"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---


---
title: "JSC370 Midterm Report"
output: html_document
date: "2025-03-17"
---

```{r setup, include=FALSE}
set.seed(1007)
library(dplyr)
library(ggplot2)
library(microbenchmark)
library(parallel)
library(rpart)
library(rpart.plot)
library(randomForest)
library(kableExtra)
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))

income <- read.csv("Data/income.csv")
cpi <- read.csv("Data/cpi.csv")
housing <- read.csv("Data/housing.csv")
```

[Link](https://github.com/tatDennis/JSC370/tree/main/Midterm) to GitHub repository.

The data set was too large to include in the GitHub repository. I've added a link
to where the data was acquired.

## Introduction

In recent years, the cost of living in Canada has risen significantly, making it
increasingly difficult for salaries to keep pace. For many young people entering
the workforce, home ownership feels less and less attainable, while those who 
already own real estate are seeing their assets appreciate in value. This has 
fueled the common belief that the older generation had a much easier time 
achieving financial stability than younger people today.

With this in mind, some interesting questions to ask are:

Has the median income of Canadians kept up with inflation?

How has the median income changed throughout the years for young and old people?
what about for men and women?

How does the change in income compare with the change in housing prices?

## The Dataset

The first data set is called 'Income of individuals by age group, sex and income 
source, Canada, provinces and selected census metropolitan areas', and contains 
detailed information about annual income from 1976 until 2022. It also divides
the data into the different provinces and territories of Canada, as well as 
different demographics. 

The second data set is called 'Consumer Price Index, annual average, 
not seasonally adjusted', and features the consumer Price index of many common
products, ranging from 1914 until 2024. The CPI is a useful statistic for 
measuring inflation, as it serves as a direct way to compare purchasing power
throughout each year.

The final data set is called 'Analytical house prices indicators', and similar
to the above consumer price index, features the change in average house price.
Although housing was contained in the above data set, it uses a different 
definition of housing which involves rent prices, whereas this analysis seeks to
understand housing ownership.

## Methods

The [income](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1110023901&pickMembers%5B0%5D=1.1&pickMembers%5B1%5D=2.5&pickMembers%5B2%5D=3.1&pickMembers%5B3%5D=4.1&cubeTimeFrame.startYear=1976&cubeTimeFrame.endYear=2022&referencePeriods=19760101%2C20220101) and [CPI](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1810000501) data sets were taken directly from the Statcan website.

[Housing price index](https://data-explorer.oecd.org/vis?lc=en&tm=DF_HOUSE_PRICES&pg=0&snb=1&vw=tb&df[ds]=dsDisseminateFinalDMZ&df[id]=DSD_AN_HOUSE_PRICES%40DF_HOUSE_PRICES&df[ag]=OECD.ECO.MPD&df[vs]=1.0&pd=1960%2C2024&dq=.A.RHP.&to[TIME_PERIOD]=false) was downloaded from OECD.


Finally, we convert each data set to use 2002 as its benchmark. This is done by
adjusting all of the data sets so that 2002 has an index value of 100, and every
other year is adjusted by this same value. 

```{r echo=FALSE}
housing <- housing |>
  mutate(Y = 100 / Value[Year == 2002],
         Adjusted.value = Value * Y) |>
  select(-Y)
```

```{r echo=FALSE}
income <- income %>%
  group_by(Age.group, Sex) %>%  
  mutate(Adjusted.value = Value / Value[Year == 2022])

income <- income %>%
  group_by(Age.group, Sex) %>%
  mutate(Y = 100 / Adjusted.value[Year == 2002], 
         Adjusted.value = Adjusted.value * Y) %>% 
  ungroup() %>%
  select(-Y)
```


We will keep these data sets separate, and merge them as needed for each question.

## Preliminary Results

### Income Index 

Lets first examine the income data set. It contains the median income of Canadians
from 1976 until 2022, and has information on male, female, and both sexes. Additionally,
it features a variety of age groups for comparison. 

An extremely interesting point to note is the years and demographics which had
the largest income index are young men aged 15-24 in the 1970s, and older 
women aged 55-64 in the 2010s and 2020s. This is most because during the 1970s,
it was more common for teenage boys to work, as well as the fact that more women
have joined the workforce compared to before.

The lowest income index is mainly comprised of older people aged 55-64 in the 1970s.
This is most likely because retirement benefits have improved by a lot since then.

Its also important to note that this does not mean these people made the most
or least amount of money, rather, the group that had the highest or lowest income
within their own group, adjusted for inflation.
```{r include=FALSE, echo=FALSE}
unique(income$Age.group)
unique(income$Sex)
summary(income)
summary(cpi)
```

```{r echo=FALSE}
kable(head(income[order(-income$Adjusted.value), ]), caption = "Largest Income Indexes", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)

kable(tail(income[order(-income$Adjusted.value), ]), caption = "Lowest Income Indexes", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)
```

### CPI

Moving on to the CPI data set, it appears that not only are cigarettes and tobacco
more expensive now than ever, but that it also had its cheapest recorded prices 
in the 1970s. The price of nicotine products has been steadily increasing in order
to encourage users to quit, which would explain the tripling in prices since 2002. 

Additionally, oil in recent years is also at an all time high, most likely due
to the ongoing conflicts around the world. 

Conversely, we see that digital computing equipment has decreased by a multiple of 30
between 1990 and 2024, as computers and phones have become more accessible in
recent years.

```{r echo=FALSE}
kable(head(cpi[order(cpi$Value), ]), caption = "Lowest CPI Values", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)

kable(tail(cpi[order(cpi$Value), ]), caption = "Largest CPI Values", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)
```

Taking a look at the overall CPI, filtering on "all items", we see that a dollar
in 1976 is worth approximately $5.17 today. Although this is more of a generalization,
it is the statistic which best captures inflation. 

```{r echo=FALSE}
cpi_filtered <- cpi %>%
  filter(Products.and.product.groups == "All-items") %>%
  filter(Year %in% c(1976, 2024))

kable(cpi_filtered, caption = "", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)

```


And just for fun, it appears that egg prices have also increased by a factor of 5. 

```{r echo=FALSE}
cpi_filtered <- cpi %>%
  filter(Products.and.product.groups == "Eggs") %>%
  filter(Year %in% c(1976, 2024))

kable(cpi_filtered, caption = "", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)
```

### Housing Index

Finally, we take a look at the housing index. 

```{r echo=FALSE}
kable(head(housing[order(housing$Adjusted.value), ]), caption = "Largest Housing Index Values", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)

kable(tail(housing[order(housing$Adjusted.value), ]), caption = "Lowest Housing Index Values", booktabs = TRUE) |>
  kable_styling(font_size = 10) |>
  kable_paper("hover", full_width = TRUE)
```

It should be no surprise that housing prices have also increased by a lot since 
the 70s, with prices reaching around recent years. Although these results appear
to be aligned with the other products earlier, the housing data takes an average
across all of Canada. Observing data sets of Vancouver or Toronto would most
likely yield more extreme results. 

## Summary

Now lets revisit some of our earlier questions from part 1. 

#### Has the median income of Canadians kept up with inflation?

Based on this output, we see that the CPI has been steadily increasing, and looks
to be accelerating in growth since 2020. However, the income index of Canadians
has mostly remained stagnant.  

```{r echo=FALSE, warning=FALSE}
both_sex <- income |>
  filter(Sex == "Both sexes") |>
  filter(Age.group == "25 to 54 years")

cpi_average <- cpi |>filter(Products.and.product.groups == "All-items") |> filter(Year >= 1976)

question1 <- merge(
  cpi_average,
  both_sex,
  by = c("Year")
)
ggplot(question1, aes(x = Year)) +
  geom_line(aes(y = Value.x, color = "CPI"), size = 1) +
  geom_line(aes(y = Adjusted.value, color = "Income Index"), size = 1, alpha = 0.7) +
  labs(title = "Consumer Pricing and Income Index Trends",
       x = "Year",
       y = "Index") +
  theme_minimal() +
  theme(legend.position = "right")
```

#### How has the median income changed throughout the years for young and old people?
What about for men and women?

We see that older Canadians have had an income increase, whereas younger Canadians
have an income decrease. Factors include less children working, seniors having 
later retirements, as well as retirement benefits in the 21st century. 

We also see that women have had an increase in their annual income, most likely
due to the fact historically, less women were part of the workforce.

```{r echo=FALSE}
both_sex_all_groups <- income |>
  filter(Sex == "Both sexes") |>
  filter(Age.group != "15 years and over")|>
  filter(Age.group != "25 to 54 years")
  
ggplot(both_sex_all_groups, aes(x = Year, y = Adjusted.value, color = Age.group)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  labs(title = "Income Index Trends by Age Group",
       x = "Year", 
       y = "Index",
       color = "Age Group") +
  theme_minimal() + 
  theme(legend.position = "bottom") 
```

```{r echo=FALSE}
male_female_income <- income |>
  filter(Sex != "Both sexes") |>
  filter(Age.group == "25 to 54 years")

ggplot(male_female_income, aes(x = Year, y = Adjusted.value, color = Sex)) +
  geom_line(size = 1) + 
  geom_point(size = 3) +
  labs(title = "Income Index Trends by Sex",
       x = "Year", 
       y = "Index",
       color = "Sex") +
  theme_minimal() + 
  theme(legend.position = "right")

```

#### How does the change in income compare with the change in housing prices?

For this question, we will look at Canadians of both sexes, aged 25-54. We see 
that the Canadian income index has not been able to keep up with the 
housing index.

```{r echo=FALSE}
question3 <- merge(
  both_sex,
  housing,
  by = c("Year")
)

ggplot(question3, aes(x = Year)) +
  geom_line(aes(y = Adjusted.value.y, color = "Housing Index"), size = 1) +
  geom_line(aes(y = Adjusted.value.x, color = "Income Index"), size = 1, alpha = 0.7) +
  labs(title = "Housing and Income Index Trends",
       x = "Year",
       y = "Index") +
  theme_minimal() +
  theme(legend.position = "right")
```




```{r}
library(tidyr)
library(dplyr)
cpi_reshape <- cpi %>%
  pivot_wider(
    names_from = Products.and.product.groups,  # New column names will come from the "Product" column
    values_from = Value    # Values will come from the "CPI" column
  )
```

```{r}
cpi_reshape_clean <- cpi_reshape |>
  select(Year, Food, Gasoline, "Mortgage interest cost", "Health care", "Dental care services",
         "Education", "Property taxes and other special charges", "Transportation", "Alcoholic beverages", "Child care services", "Toiletry items and cosmetics")

cpi_reshape_clean <- merge(cpi_reshape_clean, housing, by = "Year", all.x = TRUE)

```

```{r}
cpi_reshape_clean <- cpi_reshape_clean %>%
  rename_with(~ gsub(" ", ".", .), everything())
```


```{r}
model <- lm(Adjusted.value ~ Food + `Mortgage.interest.cost`  +
            `Health.care` + `Property.taxes.and.other.special.charges` + Year + Gasoline +
              Transportation
            , data = cpi_reshape_clean)

# View the model summary
summary(model)
summary(model)$adj.r.squared
```


```{r}
library(mgcv)

gam_model <- gam(Adjusted.value ~ s(Year) + Food + `Mortgage.interest.cost`  +
            `Health.care` + `Property.taxes.and.other.special.charges` + Gasoline +
              Transportation, data = cpi_reshape_clean)

summary(gam_model)

```

```{r}
library(caret)
project_train_index <- createDataPartition(cpi_reshape_clean$Adjusted.value, p = 0.7, list = FALSE)
project_train <- cpi_reshape_clean[project_train_index, ]
project_test <- cpi_reshape_clean[-project_train_index, ]


```

```{r}
tree <- rpart(Adjusted.value ~ ., 
                    data = project_train, 
                    method = "class",
                    control = rpart.control(minsplit = 10, 
                                            minbucket = 3, 
                                            cp = 0.01, 
                                            xval = 10))
```


```{r}
plotcp(tree)
printcp(tree)

optimalcp = tree$cptable[which.min(tree$cptable[, "xerror"]), "CP"]
```

```{r}
tree_prune <- prune(tree, cp = optimalcp)
```

```{r}
bag<- randomForest(Adjusted.value~ ., 
                         data = project_train, 
                         mtry = ncol(project_train) - 1, 
                         importance = TRUE,
                         na.action = na.omit)
# oob error rate
bag_oob <- bag$mse[length(bag$mse)]
bag_oob

importance(bag)
```

```{r}
rf <- randomForest(Adjusted.value ~ ., 
                         data = project_train, 
                         importance = TRUE, 
                         na.action = na.omit)
#oob error rate
rf_oob <- rf$mse[length(rf$mse)]
rf_oob

importance(rf)

```




```{r}
library(knitr)

model_performance <- data.frame(
  Model = c(
    "Bagging",
    "Random Forest"
  ),
  Error_MSE = c(
    bag_oob,
    rf_oob
  )
)

kable(model_performance, col.names = c("Model", "Error / MSE"), caption = "Model Performance Comparison")
```


```{r}
predictions <- predict(model)

residuals <- cpi_reshape_clean$Adjusted.value - predictions

MSE <- mean(residuals^2)
print(MSE)

gam_predictions <- predict(gam_model)

gam_residuals <- cpi_reshape_clean$Adjusted.value - gam_predictions

gam_MSE <- mean(gam_residuals^2)
print(gam_MSE)
```




